---
title: "1: R Intro and Data Visualization Challenge"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Explore components of effective data visualization
2. Apply R coding skills to create effective visualizations
3. Review and/or develop skills in R coding and syntax

## Opening Discussion

What makes an effective data visualization? What are the essential components/characteristics, and what are optional but nice-to-have components/characteristics? 


## Exploring web resources for visualization

Here are a couple of my favorites. Browse through these to get some useful ideas for R code that will improve the functionality and graphic design of your data visualization: 

[From Data to Viz](https://www.data-to-viz.com/)

[Data Visualization Example from Our Coding Club](https://ourcodingclub.github.io/tutorials/dataviz-beautification/)


## A note on color palettes

In general, the ggplot base color palettes are not the most effective option for data visualization. Compiled below is a list of color palettes available in R that may serve your purposes better. Note that when working with multiple colors, you must choose between sequential, diverging, and qualitative color palettes. Under what circumstances might each be useful?

RColorBrewer (package)

* http://colorbrewer2.org
* https://moderndata.plot.ly/create-colorful-graphs-in-r-with-rcolorbrewer-and-plotly/

viridis and viridisLite (packages)

* https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
* https://ggplot2.tidyverse.org/reference/scale_viridis.html

colorRamp (function; comes with base R as part of the grDevices package)

* https://bookdown.org/rdpeng/exdata/plotting-and-color-in-r.html#colorramp

MetBrewer: Palettes inspired by works at the Metropolitan Museum of Art in New York (package)

* https://github.com/BlakeRMills/MetBrewer


## Session Set Up

```{r}
getwd()

# install.packages("tidyverse")
# install.packages("dataRetrieval")
# install.packages("zoo")
# install.packages("ggrepel")

library(tidyverse)
library(dataRetrieval)
library(zoo)

```

## Data Visualization Challenge

We will be exploring discharge data for the Eno River, aided by the [dataRetrieval package](https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html). A basic ggplot is provided for you here. Your assignment this week will be to edit the dataset and update the graph with some data visualization best practices. 

```{r}
# Import data
EnoDischarge <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2012-01-01",
                     endDate = "2021-12-31")

# Look at the data frame in your Environment tab.

# Renaming columns (one method of multiple)
names(EnoDischarge)[4:5] <- c("Discharge_cfs", "Approval.Code")

# dataRetrieval also includes attribute information
attr(EnoDischarge, "variableInfo")
  # note: imperial, not metric
attr(EnoDischarge, "siteInfo")
  
# Build a ggplot
EnoPlot <- 
  ggplot(EnoDischarge, aes(x = Date, y = Discharge_cfs)) +
         geom_line() +
         xlab("Year")
EnoPlot
```


