---
title: "12: Mapping"
author: "Water Data Analytics | Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives

1. Delineate watersheds and flowlines using nhdplusTools
2. Extract spatial information from delineated watersheds
3. Create maps of river networks

```{r setup, include=FALSE}

library(tidyverse)
library(dataRetrieval)
library(maps)
library(sf)
# install.packages("nhdplusTools")
library(nhdplusTools)

theme_set(theme_classic())
options(scipen = 5)

```

## NHDPlus in R

The National Hydrography Dataset (NHD), now operating as NHDPlus, is a geospatial hydrology dataset developed by EPA and USGS. The `nhdplusTools` R package allows us to access NHDPlus information for use in spatial applications for rivers and streams in the US. This package works well when used in concert with the `dataRetrieval` and `sf` packages.

Key terms from the [nhdplusTools guide](https://usgs-r.github.io/nhdplusTools/): 

"**Flowline:** The NHD name for a hydrographic representation of a flowing body of water. Flowline is generally used when referring to geometry."

"**Flowpath:** The HY_Features name for a hydrologic feature that is the primary path water follows through a catchment; either from headwater to outlet or inlet to outlet. Flowpath is used when describing aspects of the abstract flowpath featuretype, generally in relation to a flowpath’s relationship to a catchment."

"**Catchment:** The most abstract unit of hydrology in HY_Features is the catchment. It is a physiographic unit with zero or one inlets and one outlet. It does not inherently have any conceptual realizations. Rather, a given catchment can be realized in a number of ways; flowpath, divide, and networks of flowpaths and divides are the primary realizations."

"**Catchment divide:** NHD “catchment” polygons are more accurately described as “catchment divide” features. Because of the overlap with the HY_Features abstract “catchment” feature type, “catchment divide” is used for polygon representations of catchments."

We will be going through a basic example today, but there are numerous extensions that can be explored. A good start guide can be found [here](https://usgs-r.github.io/nhdplusTools/articles/nhdplushr.html)

### Watershed and flowpath delineation 

Let's delineate the upstream watershed for the sampling and gage site: Neuse River at Kinston, NC.  

```{r}
# Recall there are a lot of parameters measured at the site
NeuseParams <- whatNWISdata(siteNumbers = "02089500")

# Extract latitude and longitude for the site
NeuseCoords <- NeuseParams %>%
  select(site_no, dec_lat_va, dec_long_va) %>%
  distinct()

# Define the gage site as the starting point
start_point <- st_sfc(st_point(c(NeuseCoords$dec_long_va, NeuseCoords$dec_lat_va)), 
                      crs = 4269) # NAD83, commonly  used by US agencies
start_comid <- discover_nhdplus_id(start_point)
# start_point2 <- st_as_sf(data.frame(x = NeuseCoords$dec_long_va, y =  NeuseCoords$dec_lat_va), 
#                             coords = c("x", "y"), crs = 4269)

# Navigate the NLDI network
NLDI <- navigate_nldi(list(featureSource = "comid", featureID = start_comid), 
                          mode = "upstreamTributaries", 
                          distance_km = 1000)

# Extract watershed and flowpath information
subset_file <- tempfile(fileext = ".gpkg")
subset <- subset_nhdplus(comids = as.integer(NLDI$UT$nhdplus_comid),
                         output_file = subset_file,
                         nhdplus_data = "download", 
                         flowline_only = FALSE,
                         return_data = TRUE, overwrite = TRUE)

# Create data frames
flowline <- subset$NHDFlowline_Network
catchment <- subset$CatchmentSP
waterbody <- subset$NHDWaterbody

class(flowline)
class(catchment)
class(waterbody)

# find gages near watershed
gages <- get_nwis(AOI = catchment)
class(gages)

# find gages only within watershed
gages <- st_intersection(gages, catchment)

```

### Mapping the network

The `plot_nhdplus` function is a ready-made option that includes a basemap. We can also use ggplot and the `geom_sf` function to create plots from the ground up.

Extensions: 

* [Advanced Network Attributes](https://usgs-r.github.io/nhdplusTools/articles/advanced_network.html) including hydrosequence and level path
* [Indexing and Referencing](https://usgs-r.github.io/nhdplusTools/articles/indexing.html)
* [Advanced plotting](https://usgs-r.github.io/nhdplusTools/articles/plot_nhdplus.html)


```{r}

plot_nhdplus("USGS-02089500", streamorder = 3)

states <- st_as_sf(map(database = "state", plot = FALSE, fill = TRUE, col = "white"))
nc <- filter(states, ID == "north carolina")

ggplot(nc) +
  geom_sf(fill = "white") +
  geom_sf(data = flowline, aes(color = streamorde)) +
  labs(color = "Stream Order") +
  theme(legend.position = "top")

ggplot(catchment) +
  geom_sf(fill = "white", color = "gray", lwd = 0.5) +
  geom_sf(data = flowline, aes(color = streamorde)) +
  geom_sf(data = gages, color = "darkred", size = 1) +
  labs(color = "Stream Order") +
  theme(legend.position = "top")

```

### Analyzing tabular information from the network

1. How are flowlines distributed with respect to stream order?
2. How do flowline slopes compare across stream order?
3. Total drainage area

```{r}
ggplot(flowline, aes(x = streamorde)) +
  geom_bar()

ggplot(subset(flowline, slope >= 0),
       aes(x = as.factor(streamorde), y =slope)) +
  geom_boxplot() +
  scale_y_log10()
  
max(flowline$totdasqkm)

```
