---
title: "Assignment 9: Mapping"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on mapping

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, check your PDF against the key and then submit your assignment completion survey at https://forms.gle/NDWEUu73LooFJPVM8 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-04-12

## Setup

1. Load the tidyverse, LAGOSNE, maps, dataRetrieval, sf, and nhdplusTools packages. Set your ggplot theme (can be theme_classic or something else).


```{r setup}

```

## LAGOS-NE

2. Choose five lakes in the LAGOS-NE database that are located within the same state and have chlorophyll data. Subset your data accordingly, and create two plots:

* A map of the five lakes within the state boundary, with each lake point as its own color.
* A boxplot with jittered points layered on top of chlorophyll concentrations in each lake (chlorophyll on y axis, lake on x axis), with each lake having a fill and/or color that matches the map. 

```{r}


```


## NHDPlus

3. Delineate and map the watershed and flowpaths in the network upstream of the Eno River gage in Durham (USGS-02085070). Your map should include all stream orders.


```{r}


```

4. What is the upstream drainage area for this gage, and how are the flowpaths distributed with regard to stream order?

> 

5. Are there any other gage sites in the upstream catchment? If so, where are they?
> 
